Sequence Specific Modeling of Cell-Free Protein Synthesis

### 2016-12-19 Paper_v1.tex
Inital changes

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Meta-data block
%% ---------------
%% Each author should be given as a separate \author command.
%%
%% Corresponding authors should have an e-mail given after the author
%% name as an \email command. Phone and fax numbers can be given
%% using \phone and \fax, respectively; this information is optional.
%%
%% The affiliation of authors is given after the authors; each
%% \affiliation command applies to all preceding authors not already
%% assigned an affiliation.
%%
%% The affiliation takes an option argument for the short name.  This
%% will typically be something like "University of Somewhere".
%%
%% The \altaffiliation macro should be used for new address, etc.
%% On the other hand, \alsoaffiliation is used on a per author basis
%% when authors are associated with multiple institutions.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\author{Michael Vilkhovoy}
\author{Nicholas Horvath}
\author{Jeffrey D. Varner}
\email{jdv27@cornell.edu}
\phone{+1 607 2554258}
\fax{+1 607 2559166}
\affiliation[Cornell University]
{Robert Frederick Smith School of Chemical and Biomolecular Engineering, Cornell University, Ithaca, NY 14853}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{Analysis of Cell-Free Synthetic Circuits using Sequence Specific Constraints Based Modeling}
%\abbreviations{MV,NH,JDV}
\keywords{Synthetic biology, Constraints based modeling, Biochemical modeling}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Start the main part of the manuscript here.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
Cell free systems offer many advantages for the study, manipulation and modeling of metabolism compared to \textit{in vivo} processes.
Central amongst these advantages is direct access to metabolites and the microbial biosynthetic machinery without the interference of a cell wall.
This allows us to control as well as interrogate the chemical environment while the biosynthetic machinery is operating, potentially at a fine time resolution.
Second, cell-free systems also allow us to study biological processes without the complications associated with cell growth.
Cell-free protein synthesis (CFPS) systems are arguably the most prominent examples of cell-free systems used today \cite{Jewett:2008aa}.
However, CFPS is not new; CFPS in crude \textit{E. coli} extracts has been used since the 1960s to explore fundamentally important biological mechanisms \cite{MATTHAEI:1961aa,NIRENBERG:1961aa}.
Today, cell-free systems are used in a variety of applications ranging from therapeutic protein production \cite{Lu:2014aa} to synthetic biology \cite{Hodgman:2012aa}.
Interestingly, many of the challenges confronting in-vivo genome-scale kinetic modeling can potentially be overcome in a cell-free system.
For example, there is no complex transcriptional regulation to consider, transient metabolic measurements are easier to obtain, and we no longer have to consider cell growth.
Thus, cell-free operation holds several significant advantages for model development, identification and validation.
Theoretically, genome-scale cell-free kinetic models may be possible for industrially important organisms, such as \textit{E. coli} or \textit{B. subtilis}, if a simple, tractable framework for integrating allosteric regulation with enzyme kinetics can be formulated.

@@

\section*{Results}
\subsection{Model Derivation}
The goal of this work was first to construct a modeling framework to describe cell-free protein synthesis systems and to examine its performance in productivity and yield for a protein of interest.
One mathematical framework that has found wide use in modeling metabolism are constraint based models such as flux balance analysis (FBA).
FBA can predict how cells utilize nutrients to produce products by using the biochemical stoichiometry and thermodynamical feasibility under pseudo steady-state conditions.
Traditionally, FBA is used to model \textit{in vivo} processes, however cell-free systems do not have growth associated reactions or transport through the cell membrane.
Thus, to model cell-free metabolism we constructed a cell-free stoichiometric network by removing growth associated reactions from the MG1655 reconstruction \cite{Feist:2007aa}, and incorporating amino acid synthesis and transcription/translation associated reactions \cite{Allen:2003aa} for a protein of interest to be expressed. 
The network consisted of 281 reactions and 132 species. 
We developed a cell-free sequence specific flux balance analysis (ssFBA) with a detailed promoter model \cite{Moon:2012aa} to examine the performance of CFPS. 
We first validated the ssFBA approach by comparing simulated and measured concentrations of two proteins from two different cell-free \textit{E. coli} extract systems.
The first protein, chloramphenicol acetyltransferase (CAT), was produced under a T7 promoter in a glucose/NMP cell-free system \cite{2005_calhoun_BiotechnologyProgress} for a duration of 1 hour under glucose consumption (Fig \ref{fig:CAT_GFP_prod}A).
The second protein, dual emission green fluorescent protein (deGFP), was produced under a P70 promoter in TXTL 2.0 \textit{E. coli} extract for a duration of 8 hours under maltose consumption(Fig \ref{fig:CAT_GFP_prod}B).
The ssFBA simulations predicted the production of both proteins for the duration of both CFPS batch reactions.
Uncertainty in experimental factors such as RNA polymerase, ribosome concentrations, elongation rates or the upper bounds for oxygen and glucose consumption rates did not alter the qualitative performance of the model. 
Thus, the metabolic network and molecular description of trancription and translation were consistent with experimental measurements.
\begin{figure}[t!]
\centering
\includegraphics[width=1.00\textwidth]{./Figures/CAT_GFP_prod.pdf}
\caption{Sequence specific flux balance analysis of protein production vesus time. A. CAT production under a T7 promoter in CFPS \textit{E. coli} extract for 1 hr under glucose consumption. B. deGFP production under a P70 promoter in TXTL 2.0 \textit{E. coli} extract for 8 hr under glucose consumption. 95\% CI (blue region) over the ensemble of 100 sets.}
\label{fig:CAT_GFP_prod}
\end{figure}

@@

\subsection{Examining CFPS Performance: Productivity}
Our next goal was to examine the performance of CFPS productivity for eight different proteins under three different cases. Each of the proteins were produced under a P70 promoter, except for CAT which was produced under a T7 promoter. In all our cases, CFPS is supplied with glucose. The first case, CFPS is supplied with amino acids and the system can synthesis amino acids (control). In the second case, CFPS is supplied with amino acids, however the system cannot synthesis amino acids (AA uptake w/o synthesis). We turned off these synthesis reactions because during the cell-free extract preparation the cells are often supplied with amino acids, thus the enzymes responsible for amino acid synthesis would not be present. In the third case, CFPS is not supplied with amino acids, but the system can synthesis them (AA synthesis w/o uptakte).    
We used ssFBA to estimate the productivity of eight proteins for each case (Fig \ref{fig:Prod_POI}A).
The second case (without amino acid synthesis) showed the highest productivity for each of the proteins, however the control case had very similar performance.
This shows the system has sufficient substrates to power the system and synthesis each protein. 
The third case had the lowest productivity for each protein, thus the addition of amino acids to CFPS extract is important for maintaining a relatively high productivity.
The qualitative trend of productivity for the three different cases was the same, however some proteins had higher productivity than others. 
For instance, in the second case FGF21 had a productivity of 17 ($\mu$M/h) whereas FII had a productivity of 3 ($\mu$M/h).
To examine this further, the mean productivity was plotted against the carbon number of each protein (Fig \ref{fig:Prod_POI}B).
The proteins with the highest productivity had the lowest carbon number whereas proteins with low productivity had higher carbon numbers.     
This inverse qualitative trend was due to the fact that larger proteins require more amino acids and substrates to assemble them resulting in lower productivity. 
\begin{figure}[t!]
\centering
\includegraphics[width=1.00\textwidth]{./Figures/Prod_POI.pdf}
\caption{CFPS productivity performace for eight proteins for the control (black), AA uptake w/o synthesis (off white), and AA synthesis w/o uptake (grey). A. Productivity normalized to the specific glucose uptake rate (Error bars represent the 95\% CI of the ensemble). B. Mean productivity versus the carbon number for each corresponding protein.}
\label{fig:Prod_POI}
\end{figure}

\subsection{Examining CFPS Performance: Carbon Yield}
Following the same outline as in examining the productivity, we calculated the carbon yield for each protein (Fig \ref{fig:Yield_POI}A). 
The same trends followed, where the case without amino acid synthesis showed the highest carbon yield and the control case had comparable performance. 
The third case (with no amino acid uptake) had the lowest yields; this is most likely because glucose is now utilized to synthesize the necessary amino acids for each protein as well as power the system. 
We also investigated the carbon contribution of glucose and each amino acid toward the production of deGFP for all three cases (Table \ref{tbl:yield_breakdown}).
For the first case, the system is relying on a mixture of glucose and some amino acids for the productin of deGFP with a carbon yield of 20.5\%.
Once amino acid synthesis is removed from the network (2nd case), each amino acid is utilized and the carbon yield is increased to 22.2\%.
Only the necessary amount of amino acids are used for the production of CAT, thus it may be hypothesized that all the glucose is used to power CFPS and does not contribute to the carbon yield. 
If that is the case, the carbon yield without glucose contribution is 100\%, whereas for the first case its 92.5\%.
Finally for the third case (without amino acid supplemented), the carbon yield reduces to 14.8\% and the system uses about double the amount of glucose as the first two cases. 
For this case glucose is being used to synthesize amino acids and power the system and this trend was seen across all proteins except for CAT.

@@

\subsection{Sensitivity analysis on CFPS Performance}
To better understand the effect of substrate utilization and transcription/translation parameters on CFPS performance we performed global sensitivity analysis on productivity and carbon yield for deGFP, a representative protein (Fig \ref{fig:SI_GFP}).
In examining productivity performance (Fig \ref{fig:SI_GFP}A), the significance of transcription/translation parameters were fairly constant across all three cases, with the elongation rate by ribosomes being the most sensitive.
This shows the rate of translation is instrumental in achieving high productivity, and should be the first step to investigate in order to optimize productivity, prior to examining transcription rates, this is consistent with literature findings.
Underwood and coworkers have also shown that an increase in ribosome levels does not significantly increase protein yields or rates; however, adding elongation factors increased yields by 23\% at 30 minutes \cite{2005_underwood_biotech}.
In addition, Li et al. have increased productivity of firefly lucifease by 5-fold in CFPS systems by adding and adjusting factors that affect transcription and translation such as elongation factors, ribosome recyclcing factor, release factors, chaperones, BSA, and tRNAs \cite{2014_li_PlosOne}.
In examining the substrate utilization, glucose uptake was not very important for productivity in the first two cases, but its significance increased when amino acids was removed from CFPS.
This makes sense, as amino acid synthesis can only come from glucose and becomes the only way to power protein synthesis.
On the other hand, when amino acid synthesis is removed, amino acid uptake becomes more important, as it becomes the only source of amino acids.
\begin{figure}[t!]
\includegraphics[width=1.00\textwidth]{./Figures/SI_GFP.pdf}
\caption{Total order sensitivity of deGFP productivity (A) and carbon yield (B) to specific uptake rates and transcription/translation parameters for three cases: control (black), amino acid uptake without synthesis (off white), and amino acid synthesis without uptake (gray). (Error bars represent 95\% CI of the ensemble).}
\label{fig:SI_GFP}
\end{figure}

@@

\section*{Materials and Methods}
\subsection*{Formulation and solution of the model equations.}
The flux balance analysis problem was formulated as:
\begin{equation}\nonumber
 \begin{multlined}
	\qquad \qquad \qquad \max_{\boldsymbol{w}}{} \! \left( w_{obj} = \mathbf{\boldsymbol{\theta}}^T \boldsymbol{w} \right) \\
	\mathrm{Subject \; to:}
	 \; \; \mathbf{S}\mathbf{w}=\mathbf{0} \\
\alpha_i \leq w_i \leq \beta_i  \qquad i=1,2,\hdots,\mathcal{R}
 \end{multlined}
\end{equation}
where $\mathbf{S}$ denotes the stoichiometric matrix, $\mathbf{w}$ denotes the unknown flux vector, $\boldsymbol{\theta}$ denotes the objective selection vector
and $\alpha_i$ and $\beta_i$ denote the lower and upper bounds on flux $w_{i}$, respectively.
The flux balance analysis problem was solved using the GNU Linear Programming Kit (v4.52) \cite{GLPK}.

\subsection*{Sequence-specific flux balance analysis}
We estimated the theoretical maximum carbon yield using sequence-specific flux balance analysis (ssFBA) \cite{Allen:2003aa}.
The sequence-specific flux balance analysis problem was formulated as a linear program:
\begin{equation}
 \begin{multlined}
	\qquad \qquad \qquad \max_{\boldsymbol{w}}{} \! \left( w_{TL} = \mathbf{\boldsymbol{\theta}}^T \boldsymbol{w} \right) \\
	\mathrm{Subject \; to:}
	 \; \; \mathbf{S}\mathbf{w}=\mathbf{0} \\
\alpha_i \leq w_i \leq \beta_i  \qquad i=1,2,\hdots,\mathcal{R}
 \end{multlined}
\end{equation}
where $\mathbf{S}$ denotes the stoichiometric matrix, $\mathbf{w}$ denotes the unknown flux vector, $\boldsymbol{\theta}$ denotes the objective selection vector
and $\alpha_i$ and $\beta_i$ denote the lower and upper bounds on flux $w_{i}$, respectively.
The stoichiometry of the kinetic model was used for the ssFBA calculations, with the execpetion of the transcription and translation rates.
The transcription (TX) and translation (TL) stoichiometry was modeled using the template reactions taken from Allen and Palsson \cite{Allen:2003aa}:
\begin{eqnarray*}
\mathrm{{G}_{\mathcal{P}}}+\mathrm{{R}_{1}} &\longrightarrow& \mathrm{G_{\mathcal{P}}^{*}} \\
\mathrm{G_{\mathcal{P}}^{*}}+\sum_{\mathrm{k\in\left\{A,C,G,U\right\}}}\eta_{k}\cdot \mathrm{\left\{k\right\}TP} &\overset{\rm TX}{\longrightarrow}& \mathrm{mRNA}+\mathrm{G_{\mathcal{P}}}+\mathrm{R_{1}}+ \mathrm{\sum_{k\in\left\{A,C,G,U\right\}}2\eta_{k}\cdot Pi}\\
\mathrm{mRNA} &\longrightarrow& \mathrm{\sum_{k\in\left\{A,C,G,U\right\}}\eta_{k}\cdot \left\{k\right\}MP} \\
\mathrm{mRNA}+\mathrm{R_{2}} &\longrightarrow& \mathrm{R_{2}^{*}} \\
\mathrm{\alpha_{j}\cdot AA_{j}+\alpha_{j}\cdot tRNA+\alpha_{j}\cdot ATP} &\longrightarrow& \mathrm{\alpha_{j}\cdot AA_{j}-tRNA_{j}+}\\
&& \mathrm{\qquad \alpha_{j}\cdot AMP+2\alpha_{j}\cdot Pi} \qquad{j=1,2,\hdots,20}\\
\mathrm{R_{2}^{*}+\sum_{j\in\left\{AA\right\}}\alpha_{j}\cdot \Big(AA_{j}-tRNA_{j}+2\cdot GTP\Big)} &\stackrel{\rm TL}{\longrightarrow}& \mathcal{P}+\mathrm{R_{2}+mRNA}+\\
&& \mathrm{\quad +\sum_{j\in\left\{AA\right\}}\alpha_{j}\cdot\Big(tRNA+2\cdot GDP+2\cdot Pi\Big)}
\end{eqnarray*}
where $G_{\mathcal{P}}$ denotes the gene encoding protein product $\mathcal{P}$,
$\rm R_{1}$ denotes the concentration of RNA polymerase,
$G_{\mathcal{P}}^{*}$ denotes the gene bounded by the RNA polymerase,
$\eta_{i}$ and $ \alpha_{j}$ denote the stoichiometric coefficients for nucleotide and amino acid, respectively,
$\rm Pi$ denotes inorganic phosphate,
$\rm R_{2}$ denotes the ribosome concentration,
$\rm R_{2}^{*}$ denotes bounded ribosome,
and $AA_{j}$ denotes $j^{th}$ amino acid.

The transcription rate ($w_{TX}$) was fixed in the ssFBA calculation at:
\begin{equation}
	w_{TX} = V_{TX}^{max}\left(\frac{G}{K_{TX}+G}\right)
\end{equation}
where $G$ denotes the gene concentration, and $K_{TX}$ denotes a transcription saturation coefficient.
The maximum rate of transcription $V_{TX}^{max}$ was formulated as:
\begin{equation}
	V_{TX}^{max} \equiv \left[R_{1}\left(\frac{v_{TX}}{l_{G}}\right)P\right]
\end{equation}
The term $R_{1}$ denotes the RNA polymerase abundance,
$v_{TX}$ denotes the RNA polymerase elongation rate (nt/hr),
$l_{G}$ denotes the gene length in nucleotides, and $P$ describes the promoter activity.
 The T7 binding promoter was used for CAT production was formulated as:
\begin{equation}
	P_{T7} = (\frac{K_{T7}}{1 + K_{T7}}
\end{equation}
where $K_{T7}$ denotes a T7 RNA polymerase binding constant \cite{Moon:2012aa}.
The $\sigma_{70}$ binding promoter was used for all other proteins was formulated as:
\begin{equation}
	P_{\sigma_{70}} = \frac{K_{1}+K_{2}f_{p70}}{1 + K_{1}+K_{2}f_{p70}}
\end{equation}
where $K_{1}$ denotes the state of RNA polymerase binding, $K_{2}$ is the state of sigma-70 binding along with RNA polymerase, and $f_{p70}$ is the fraction of the transcription factor, sigma-70, bound to the promoter following Hills kinetics.

@@

The objective of the sequence-specific flux balance calculation was to maximize the rate of CAT translation, $w_{TL}$.
The total glucose uptake rate was bounded by [0,40 mM/h] according to experimental data;
while the amino acid uptake rates were bounded by [0,30 mM/h], but did not reach the maximum flux.
The protein gene and protein sequences were taken from literature and are available in the Supporting Information.
The sequence-specific flux balance linear program was solved using the GNU Linear Programming Kit (GLPK) v4.52 \cite{GLPK}.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Meta-data block
%% ---------------
%% Each author should be given as a separate \author command.
%%
%% Corresponding authors should have an e-mail given after the author
%% name as an \email command. Phone and fax numbers can be given
%% using \phone and \fax, respectively; this information is optional.
%%
%% The affiliation of authors is given after the authors; each
%% \affiliation command applies to all preceding authors not already
%% assigned an affiliation.
%%
%% The affiliation takes an option argument for the short name.  This
%% will typically be something like "University of Somewhere".
%%
%% The \altaffiliation macro should be used for new address, etc.
%% On the other hand, \alsoaffiliation is used on a per author basis
%% when authors are associated with multiple institutions.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\author{Michael Vilkhovoy}
\author{Nicholas Horvath}
\author{Jeffrey D. Varner}
\email{jdv27@cornell.edu}
\phone{+1~(607)~255-4258}
\fax{+1~(607)~255-9166}
\affiliation[Cornell University]
{Robert Frederick Smith School of Chemical and Biomolecular Engineering, Cornell University, Ithaca, NY 14853}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{Sequence Specific Modeling of Cell-Free Protein Synthesis}
%\abbreviations{MV,NH,JDV}
\keywords{Synthetic biology, Constraints based modeling, Biochemical modeling}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Start the main part of the manuscript here.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
Cell free systems offer many advantages for the study, manipulation and modeling of metabolism compared to \textit{in vivo} processes.
Central amongst these advantages is direct access to metabolites and the microbial biosynthetic machinery without the interference of a cell wall.
This allows us to control as well as interrogate the chemical environment while the biosynthetic machinery is operating, potentially at a fine time resolution.
Second, cell-free systems also allow us to study biological processes without the complications associated with cell growth.
Cell-free protein synthesis (CFPS) systems are arguably the most prominent examples of cell-free systems used today \cite{Jewett:2008aa}.
However, CFPS is not new; CFPS in crude \textit{E.~coli} extracts has been used since the 1960s to explore fundamentally important biological mechanisms \cite{MATTHAEI:1961aa,NIRENBERG:1961aa}.
Today, cell-free systems are used in a variety of applications ranging from therapeutic protein production \cite{Lu:2014aa} to synthetic biology \cite{Hodgman:2012aa}.
Interestingly, many of the challenges confronting in-vivo genome-scale kinetic modeling can potentially be overcome in a cell-free system.
For example, there is no complex transcriptional regulation to consider, transient metabolic measurements are easier to obtain, and we no longer have to consider cell growth.
Thus, cell-free operation holds several significant advantages for model development, identification and validation.
Theoretically, genome-scale cell-free kinetic models may be possible for industrially important organisms, such as \textit{E. coli} or \textit{B. subtilis}, if a simple, tractable framework for integrating allosteric regulation with enzyme kinetics can be formulated.

@@

\subsection{Model Derivation}
The goal of this work was first to construct a modeling framework to describe cell-free protein synthesis systems and to examine its performance in productivity and yield for a protein of interest.
One mathematical framework that has found wide use in modeling metabolism are constraint based models such as flux balance analysis (FBA).
FBA can predict how cells utilize nutrients to produce products by using the biochemical stoichiometry and thermodynamical feasibility under pseudo steady-state conditions.
Traditionally, FBA is used to model \textit{in vivo} processes, however cell-free systems do not have growth associated reactions or transport through the cell membrane.
Thus, to model cell-free metabolism we constructed a cell-free stoichiometric network by removing growth associated reactions from the MG1655 reconstruction \cite{Feist:2007aa}, and incorporating amino acid synthesis and transcription/translation associated reactions \cite{Allen:2003aa} for a protein of interest to be expressed.
The network consisted of 281 reactions and 132 species.
We developed a cell-free sequence specific flux balance analysis (ssFBA) with a detailed promoter model \cite{Moon:2012aa} to examine the performance of CFPS.
We first validated the ssFBA approach by comparing simulated and measured concentrations of two proteins from two different cell-free \textit{E. coli} extract systems.
The first protein, chloramphenicol acetyltransferase (CAT), was produced under a T7 promoter in a glucose/NMP cell-free system \cite{2005_calhoun_BiotechnologyProgress} for a duration of 1 hour under glucose consumption (Fig \ref{fig:CAT_GFP_prod}A).
The second protein, dual emission green fluorescent protein (deGFP), was produced under a P70 promoter in TXTL 2.0 \textit{E. coli} extract for a duration of 8 hours under maltose consumption(Fig \ref{fig:CAT_GFP_prod}B).
The ssFBA simulations predicted the production of both proteins for the duration of both CFPS batch reactions.
Uncertainty in experimental factors such as RNA polymerase, ribosome concentrations, elongation rates or the upper bounds for oxygen and glucose consumption rates did not alter the qualitative performance of the model.
Thus, the metabolic network and molecular description of trancription and translation were consistent with experimental measurements.
\begin{figure}[t!]
\centering
\includegraphics[width=1.00\textwidth]{./Figures/CAT_GFP_prod.pdf}
\caption{Sequence specific flux balance analysis of protein production vesus time. A. CAT production under a T7 promoter in CFPS \textit{E. coli} extract for 1 hr under glucose consumption. B. deGFP production under a P70 promoter in TXTL 2.0 \textit{E. coli} extract for 8 hr under glucose consumption. 95\% CI (blue region) over the ensemble of 100 sets.}
\label{fig:CAT_GFP_prod}
\end{figure}

@@

\subsection{CFPS Productivity}
Our next goal was to examine the performance of CFPS productivity for eight different proteins under three different cases. Each of the proteins were produced under a P70 promoter, except for CAT which was produced under a T7 promoter. In all our cases, CFPS is supplied with glucose. The first case, CFPS is supplied with amino acids and the system can synthesis amino acids (control). In the second case, CFPS is supplied with amino acids, however the system cannot synthesis amino acids (AA uptake w/o synthesis). We turned off these synthesis reactions because during the cell-free extract preparation the cells are often supplied with amino acids, thus the enzymes responsible for amino acid synthesis would not be present. In the third case, CFPS is not supplied with amino acids, but the system can synthesis them (AA synthesis w/o uptakte).
We used ssFBA to estimate the productivity of eight proteins for each case (Fig \ref{fig:Prod_POI}A).
The second case (without amino acid synthesis) showed the highest productivity for each of the proteins, however the control case had very similar performance.
This shows the system has sufficient substrates to power the system and synthesis each protein.
The third case had the lowest productivity for each protein, thus the addition of amino acids to CFPS extract is important for maintaining a relatively high productivity.
The qualitative trend of productivity for the three different cases was the same, however some proteins had higher productivity than others.
For instance, in the second case FGF21 had a productivity of 17 ($\mu$M/h) whereas FII had a productivity of 3 ($\mu$M/h).
To examine this further, the mean productivity was plotted against the carbon number of each protein (Fig \ref{fig:Prod_POI}B).
The proteins with the highest productivity had the lowest carbon number whereas proteins with low productivity had higher carbon numbers.
This inverse qualitative trend was due to the fact that larger proteins require more amino acids and substrates to assemble them resulting in lower productivity.
\begin{figure}[t!]
\centering
\includegraphics[width=1.00\textwidth]{./Figures/Prod_POI.pdf}
\caption{CFPS productivity performace for eight proteins for the control (black), AA uptake w/o synthesis (off white), and AA synthesis w/o uptake (grey). A. Productivity normalized to the specific glucose uptake rate (Error bars represent the 95\% CI of the ensemble). B. Mean productivity versus the carbon number for each corresponding protein.}
\label{fig:Prod_POI}
\end{figure}

\subsection{CFPS Carbon Yield}
Following the same outline as in examining the productivity, we calculated the carbon yield for each protein (Fig \ref{fig:Yield_POI}A).
The same trends followed, where the case without amino acid synthesis showed the highest carbon yield and the control case had comparable performance.
The third case (with no amino acid uptake) had the lowest yields; this is most likely because glucose is now utilized to synthesize the necessary amino acids for each protein as well as power the system.
We also investigated the carbon contribution of glucose and each amino acid toward the production of deGFP for all three cases (Table \ref{tbl:yield_breakdown}).
For the first case, the system is relying on a mixture of glucose and some amino acids for the productin of deGFP with a carbon yield of 20.5\%.
Once amino acid synthesis is removed from the network (2nd case), each amino acid is utilized and the carbon yield is increased to 22.2\%.
Only the necessary amount of amino acids are used for the production of CAT, thus it may be hypothesized that all the glucose is used to power CFPS and does not contribute to the carbon yield.
If that is the case, the carbon yield without glucose contribution is 100\%, whereas for the first case its 92.5\%.
Finally for the third case (without amino acid supplemented), the carbon yield reduces to 14.8\% and the system uses about double the amount of glucose as the first two cases.
For this case glucose is being used to synthesize amino acids and power the system and this trend was seen across all proteins except for CAT.

@@

\subsection{Sensitivity analysis on CFPS performance}
To better understand the effect of substrate utilization and transcription/translation parameters on CFPS performance we performed global sensitivity analysis on productivity and carbon yield for deGFP, a representative protein (Fig \ref{fig:SI_GFP}).
In examining productivity performance (Fig \ref{fig:SI_GFP}A), the significance of transcription/translation parameters were fairly constant across all three cases, with the elongation rate by ribosomes being the most sensitive.
This shows the rate of translation is instrumental in achieving high productivity, and should be the first step to investigate in order to optimize productivity, prior to examining transcription rates, this is consistent with literature findings.
Underwood and coworkers have also shown that an increase in ribosome levels does not significantly increase protein yields or rates; however, adding elongation factors increased yields by 23\% at 30 minutes \cite{2005_underwood_biotech}.
In addition, Li et al. have increased productivity of firefly lucifease by 5-fold in CFPS systems by adding and adjusting factors that affect transcription and translation such as elongation factors, ribosome recyclcing factor, release factors, chaperones, BSA, and tRNAs \cite{2014_li_PlosOne}.
In examining the substrate utilization, glucose uptake was not very important for productivity in the first two cases, but its significance increased when amino acids was removed from CFPS.
This makes sense, as amino acid synthesis can only come from glucose and becomes the only way to power protein synthesis.
On the other hand, when amino acid synthesis is removed, amino acid uptake becomes more important, as it becomes the only source of amino acids.
\begin{figure}[t!]
\includegraphics[width=1.00\textwidth]{./Figures/SI_GFP.pdf}
\caption{Total order sensitivity of deGFP productivity (A) and carbon yield (B) to specific uptake rates and transcription/translation parameters for three cases: control (black), amino acid uptake without synthesis (off white), and amino acid synthesis without uptake (gray). (Error bars represent 95\% CI of the ensemble).}
\label{fig:SI_GFP}
\end{figure}

@@

\section*{Materials and Methods}

\subsection*{Formulation and solution of the model equations.}
We estimated the theoretical maximum performance of the cell free protein synthesis system using sequence-specific flux balance analysis (ssFBA) \cite{Allen:2003aa}.
The sequence-specific flux balance analysis problem was formulated as a linear program:
\begin{equation}
 \begin{multlined}
	\qquad \qquad \qquad \max_{\boldsymbol{w}}{} \! \left( w_{TL} = \mathbf{\boldsymbol{\theta}}^T \boldsymbol{w} \right) \\
	\mathrm{Subject \; to:}
	 \; \; \mathbf{S}\mathbf{w}=\mathbf{0} \\
\alpha_i \leq w_i \leq \beta_i  \qquad i=1,2,\hdots,\mathcal{R}
 \end{multlined}
\end{equation}
where $\mathbf{S}$ denotes the stoichiometric matrix, $\mathbf{w}$ denotes the unknown flux vector, $\boldsymbol{\theta}$ denotes the objective selection vector
and $\alpha_i$ and $\beta_i$ denote the lower and upper bounds on flux $w_{i}$, respectively.
The stoichiometry of the kinetic model was used for the ssFBA calculations, with the execpetion of the transcription and translation rates.
The transcription (TX) and translation (TL) stoichiometry was modeled using the template reactions taken from Allen and Palsson \cite{Allen:2003aa}:
\begin{eqnarray*}
\mathrm{{G}_{\mathcal{P}}}+\mathrm{{R}_{1}} &\longrightarrow& \mathrm{G_{\mathcal{P}}^{*}} \\
\mathrm{G_{\mathcal{P}}^{*}}+\sum_{\mathrm{k\in\left\{A,C,G,U\right\}}}\eta_{k}\cdot \mathrm{\left\{k\right\}TP} &\overset{\rm TX}{\longrightarrow}& \mathrm{mRNA}+\mathrm{G_{\mathcal{P}}}+\mathrm{R_{1}}+ \mathrm{\sum_{k\in\left\{A,C,G,U\right\}}2\eta_{k}\cdot Pi}\\
\mathrm{mRNA} &\longrightarrow& \mathrm{\sum_{k\in\left\{A,C,G,U\right\}}\eta_{k}\cdot \left\{k\right\}MP} \\
\mathrm{mRNA}+\mathrm{R_{2}} &\longrightarrow& \mathrm{R_{2}^{*}} \\
\mathrm{\alpha_{j}\cdot AA_{j}+\alpha_{j}\cdot tRNA+\alpha_{j}\cdot ATP} &\longrightarrow& \mathrm{\alpha_{j}\cdot AA_{j}-tRNA_{j}+}\\
&& \mathrm{\qquad \alpha_{j}\cdot AMP+2\alpha_{j}\cdot Pi} \qquad{j=1,2,\hdots,20}\\
\mathrm{R_{2}^{*}+\sum_{j\in\left\{AA\right\}}\alpha_{j}\cdot \Big(AA_{j}-tRNA_{j}+2\cdot GTP\Big)} &\stackrel{\rm TL}{\longrightarrow}& \mathcal{P}+\mathrm{R_{2}+mRNA}+\\
&& \mathrm{\quad +\sum_{j\in\left\{AA\right\}}\alpha_{j}\cdot\Big(tRNA+2\cdot GDP+2\cdot Pi\Big)}
\end{eqnarray*}
where $G_{\mathcal{P}}$ denotes the gene encoding protein product $\mathcal{P}$,
$\rm R_{1}$ denotes the concentration of RNA polymerase,
$G_{\mathcal{P}}^{*}$ denotes the gene bounded by the RNA polymerase,
$\eta_{i}$ and $ \alpha_{j}$ denote the stoichiometric coefficients for nucleotide and amino acid, respectively,
$\rm Pi$ denotes inorganic phosphate,
$\rm R_{2}$ denotes the ribosome concentration,
$\rm R_{2}^{*}$ denotes bounded ribosome,
and $AA_{j}$ denotes $j^{th}$ amino acid.

The transcription rate ($w_{TX}$) was fixed in the ssFBA calculation at:
\begin{equation}
	w_{TX} = V_{TX}^{max}\left(\frac{G}{K_{TX}+G}\right)
\end{equation}
where $G$ denotes the gene concentration, and $K_{TX}$ denotes a transcription saturation coefficient.
The maximum rate of transcription $V_{TX}^{max}$ was formulated as:
\begin{equation}
	V_{TX}^{max} \equiv \left[R_{1}\left(\frac{v_{TX}}{l_{G}}\right)\mathcal{P}\right]
\end{equation}
The term $R_{1}$ denotes the RNA polymerase abundance,
$v_{TX}$ denotes the RNA polymerase elongation rate (nt/hr),
$l_{G}$ denotes the gene length in nucleotides, and the last term $\mathcal{P}$ describes a model of promoter activity.
In this study, we considered two promoters; the T7 and $\sigma_{70}$ promoters.
The promoter function for the T7 promoter, $\mathcal{P}_{T7}$ was given by:
\begin{equation}
	P_{T7} = (\frac{K_{T7}}{1 + K_{T7}}
\end{equation}
where $K_{T7}$ denotes a T7 RNA polymerase binding constant \cite{Moon:2012aa}.
The $\sigma_{70}$ binding promoter was used for all other proteins was formulated as:
\begin{equation}
	P_{\sigma_{70}} = \frac{K_{1}+K_{2}f_{p70}}{1 + K_{1}+K_{2}f_{p70}}
\end{equation}
where $K_{1}$ denotes the state of RNA polymerase binding,
$K_{2}$ is the state of $\sigma_{70} $binding along with RNA polymerase,
and $f_{p70}$ denotes the fraction of the $\sigma_{70}$ transcription factor bound to the promoter (modeled as a Hill function).

@@

The objective of the sequence-specific flux balance calculation was to maximize the rate of protein translation, $w_{TL}$.
The total glucose uptake rate was bounded by [0,40 mM/h] according to experimental data;
while the amino acid uptake rates were bounded by [0,30 mM/h], but did not reach the maximum flux.
The protein gene and protein sequences were taken from literature and are available in the Supporting Information.
The sequence-specific flux balance linear program was solved using the GNU Linear Programming Kit (GLPK) v4.52 \cite{GLPK}.

% The flux balance analysis problem was formulated as:
% \begin{equation}\nonumber
